

---
title: ""
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    self_contained: true
    includes:
      before_body: logo.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 150)

library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(forcats)
library(lubridate)
library(zoo)
library(plotly)
```

# Executive Summary

This report analyzes global patterns of armed conflict and civil unrest using data from the
:contentReference[oaicite:1]{index=1}
covering the period up to **February 2026**.

The objective is to provide a data-driven overview of where violence occurs, how it evolves over time,
and which regions and countries are most affected.


```{r load-data, cache=TRUE}
files <- c(
  "Africa_aggregated_data_up_to-2026-02-28.xlsx",
  "Asia-Pacific_aggregated_data_up_to-2026-02-28.xlsx",
  "Europe-Central-Asia_aggregated_data_up_to-2026-02-28.xlsx",
  "Latin-America-the-Caribbean_aggregated_data_up_to-2026-02-28.xlsx",
  "Middle-East_aggregated_data_up_to-2026-02-28.xlsx",
  "US-and-Canada_aggregated_data_up_to-2026-02-28.xlsx"
)

combined <- lapply(files, read_xlsx) |> bind_rows()

combined <- combined |>
  mutate(
    WEEK = as.Date(WEEK),
    YEAR = year(WEEK),
    BROAD_REGION = case_when(
      REGION %in% c("Northern Africa", "Eastern Africa", "Western Africa",
                    "Middle Africa",   "Southern Africa")        ~ "Africa",
      REGION %in% c("South Asia", "Southeast Asia",
                    "East Asia", "Oceania")                      ~ "Asia-Pacific",
      REGION %in% c("Caucasus and Central Asia", "Europe")       ~ "Europe & C. Asia",
      REGION %in% c("South America", "Central America",
                    "Caribbean")                                  ~ "Latin America",
      REGION == "Middle East"                                     ~ "Middle East",
      REGION == "North America"                                   ~ "N. America",
      TRUE                                                        ~ "Other"
    )
  )

civ_fat <- read_xlsx("number_of_reported_civilian_fatalities_by_country-year_as-of-06Mar2026.xlsx")
tot_fat <- read_xlsx("number_of_reported_fatalities_by_country-year_as-of-06Mar2026.xlsx")
```


# Global Trends in Conflict Events and Fatalities

```{r plot1-global-trend}
yearly <- combined |>
  filter(YEAR >= 2010, YEAR <= 2025) |>
  group_by(YEAR) |>
  summarise(
    Events     = sum(EVENTS,     na.rm = TRUE),
    Fatalities = sum(FATALITIES, na.rm = TRUE),
    .groups    = "drop"
  )

scale_factor <- max(yearly$Events) / max(yearly$Fatalities)

ggplot(yearly, aes(x = YEAR)) +
  geom_col(aes(y = Events), fill = "#2980B9", alpha = 0.8, width = 0.7) +
  geom_line(aes(y = Fatalities * scale_factor),
            colour = "#C0392B", linewidth = 1.3) +
  geom_point(aes(y = Fatalities * scale_factor),
             colour = "#C0392B", size = 3, shape = 21,
             fill = "white", stroke = 1.8) +
  scale_x_continuous(breaks = 2010:2025) +
  scale_y_continuous(
    name     = "Events recorded",
    labels   = comma,
    sec.axis = sec_axis(~ . / scale_factor, name = "Reported fatalities", labels = comma)
  ) +
  labs(
    title    = NULL,
    subtitle = "Bars = events (left axis)  ·  Line = fatalities (right axis)",
    x        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", colour = "#1a2744"),
    plot.subtitle      = element_text(colour = "#666", size = 11),
    axis.title.y.left  = element_text(colour = "#2980B9"),
    axis.title.y.right = element_text(colour = "#C0392B"),
    axis.text.x        = element_text(angle = 45, hjust = 1),
    panel.grid.minor   = element_blank()
  )
```

Over the past fifteen years, both the number of conflict events and total fatalities have increased significantly. 
This indicates that global instability is becoming more frequent and more lethal.

# Global Distribution of Conflict Fatalities

```{r plot-world-heatmap, fig.width=5, fig.height=2}
library(maps)
library(ggplot2)
library(dplyr)
library(scales)
library(plotly)

world_map <- map_data("world") |>
  filter(lat > -60)

country_fat_total <- tot_fat |>
  filter(YEAR < 2026) |>
  group_by(COUNTRY) |>
  summarise(total = sum(FATALITIES, na.rm = TRUE), .groups = "drop") |>
  mutate(
    COUNTRY = recode(COUNTRY,
      "United States" = "USA",
      "United Kingdom" = "UK",
      "Democratic Republic of Congo" = "Democratic Republic of the Congo",
      "Republic of Congo" = "Republic of Congo",
      "Czechia" = "Czech Republic",
      "North Macedonia" = "Macedonia",
      "Eswatini" = "Swaziland",
      "Trinidad and Tobago" = "Trinidad",
      "Antigua and Barbuda" = "Antigua",
      "Saint Kitts and Nevis" = "Saint Kitts",
      "Palestine" = "Palestine"
    ),
    fatality_group = cut(
      total,
      breaks = c(0, 1000, 10000, 50000, 100000, Inf),
      labels = c("Low", "Moderate", "High", "Very High", "Extreme"),
      include.lowest = TRUE
    )
  )

map_joined <- world_map |>
  left_join(country_fat_total, by = c("region" = "COUNTRY"))

p <- ggplot(map_joined,
            aes(
              long, lat,
              group = group,
              fill = fatality_group,
              text = paste0(
                "Country: ", region,
                "<br>Fatalities: ", comma(total)
              )
            )) +

  geom_polygon(colour = "white", linewidth = 0.15) +

  scale_fill_manual(
    values = c(
      "Low" = "#fde0dd",
      "Moderate" = "#fcbba1",
      "High" = "#fb6a4a",
      "Very High" = "#cb181d",
      "Extreme" = "#67000d"
    ),
    na.value = "#d9d9d9",
    name = "Fatality concentration"
  ) +

  coord_quickmap(expand = FALSE) +

  theme_void()

# convert to interactive plot
ggplotly(p, tooltip = "text") %>%
  layout(
    margin = list(l = 0, r = 0, t = 0, b = 0), height = 340
  ) %>%
  config(displayModeBar = FALSE)
```
This map highlights the unequal global distribution of conflict-related fatalities. 
A small number of countries account for a disproportionately large share of reported deaths, revealing concentrated conflict hotspots.

# Global Distribution of Conflict Fatalities

```{r plot-fatalities-by-region-pie}
library(dplyr)
library(ggplot2)
library(scales)

region_fatalities <- combined |>
  filter(YEAR >= 2016) |>
  group_by(BROAD_REGION) |>
  summarise(fatalities = sum(FATALITIES, na.rm = TRUE), .groups = "drop") |>
  mutate(share = fatalities / sum(fatalities))

ggplot(region_fatalities, aes(x = "", y = share, fill = BROAD_REGION)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  scale_y_continuous(labels = percent) +
  labs(
    subtitle = "Share of total reported deaths by region (2016–2025)",
    fill = NULL
  ) +
  theme_void()
```

This chart shows how conflict-related fatalities are distributed across global regions. It highlights that a small number of regions account for most of the reported deaths, emphasizing the strong geographic concentration of global conflict.

# Types of Conflict Events by Region 

```{r plot-region-composition}
event_pal <- c(
  "Battles"                    = "#C0392B",
  "Explosions/Remote violence" = "#E67E22",
  "Violence against civilians" = "#8E44AD",
  "Protests"                   = "#2980B9",
  "Riots"                      = "#16A085",
  "Strategic developments"     = "#95A5A6"
)

region_comp <- combined |>
  filter(BROAD_REGION != "Other", YEAR >= 2016) |>
  group_by(BROAD_REGION, EVENT_TYPE) |>
  summarise(Events = sum(EVENTS, na.rm = TRUE), .groups = "drop") |>
  group_by(BROAD_REGION) |>
  mutate(
    Share        = Events / sum(Events),
    BROAD_REGION = fct_reorder(BROAD_REGION, Events, .fun = sum, .desc = FALSE)
  ) |>
  ungroup() |>
  mutate(EVENT_TYPE = factor(EVENT_TYPE, levels = names(event_pal)))

ggplot(region_comp, aes(x = BROAD_REGION, y = Share, fill = EVENT_TYPE)) +
  geom_col(width = 0.7, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = event_pal, name = NULL) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = c(0, 0)) +
  coord_flip() +
  labs(
    title    = NULL,
    subtitle = "Each bar sums to 100% - shows the dominant mode of conflict per region",
    x        = NULL,
    y        = "Share of events",
    caption  = "Source: ACLED"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold", colour = "#1a2744"),
    plot.subtitle      = element_text(colour = "#666", size = 11),
    plot.caption       = element_text(colour = "#999", size = 9),
    legend.position    = "bottom",
    legend.key.size    = unit(0.4, "cm"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  ) +
  guides(fill = guide_legend(nrow = 2))
```

Conflict manifests differently across regions. 
While some regions are dominated by armed battles, others experience more protests, riots, or violence against civilians, reflecting diverse political and social dynamics.

# Top 15 Countries by Conflict Fatalities

```{r plot-top15}
top15 <- tot_fat |>
  filter(YEAR < 2026) |>
  group_by(COUNTRY) |>
  summarise(total = sum(FATALITIES, na.rm = TRUE), .groups = "drop") |>
  slice_max(total, n = 15) |>
  left_join(
    civ_fat |>
      filter(YEAR < 2026) |>
      group_by(COUNTRY) |>
      summarise(civilian = sum(FATALITIES, na.rm = TRUE), .groups = "drop"),
    by = "COUNTRY"
  ) |>
  mutate(
    non_civilian = total - civilian,
    COUNTRY      = fct_reorder(COUNTRY, total)
  ) |>
  pivot_longer(cols = c(civilian, non_civilian),
               names_to = "type", values_to = "fatalities") |>
  mutate(type = recode(type,
    "civilian"     = "Civilian",
    "non_civilian" = "Non-civilian"
  ))

ggplot(top15, aes(x = COUNTRY, y = fatalities, fill = type)) +
  geom_col(width = 0.7) +
  scale_fill_manual(
    values = c("Civilian" = "#C0392B", "Non-civilian" = "#BDC3C7"),
    name   = NULL
  ) +
  scale_y_continuous(labels = comma) +
  coord_flip() +
  labs(
    title    = NULL,
    subtitle = "Red = civilian fatalities · grey = non-civilian",
    x        = NULL,
    y        = "Reported fatalities",
    caption  = "Source: ACLED"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.subtitle      = element_text(colour = "#666", size = 11),
    plot.caption       = element_text(colour = "#999", size = 9),
    legend.position    = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

The countries with the highest total fatalities are not always those with the highest civilian losses. 
This comparison helps distinguish conflicts primarily affecting combatants from those heavily impacting civilian populations.

# Ukraine Conflict Timeline

```{r plot-ukraine-weekly}
ukraine_weekly <- combined |>
  filter(COUNTRY == "Ukraine", YEAR >= 2020) |>
  group_by(WEEK) |>
  summarise(
    Fatalities = sum(FATALITIES, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  arrange(WEEK) |>
  mutate(rolling_avg = rollmean(Fatalities, k = 4, fill = NA, align = "right"))

ggplot(ukraine_weekly, aes(x = WEEK)) +
  geom_col(aes(y = Fatalities), fill = "#C0392B", alpha = 0.4, width = 5) +
  geom_line(aes(y = rolling_avg), colour = "#C0392B", linewidth = 1.2) +
  geom_vline(xintercept = as.Date("2022-02-24"),
             linetype = "dashed", colour = "#1a2744", linewidth = 0.8) +
  annotate("text",
           x     = as.Date("2022-05-01"),
           y     = max(ukraine_weekly$Fatalities, na.rm = TRUE) * 0.95,
           label = "Full-scale invasion\n24 Feb 2022",
           hjust = 0, colour = "#1a2744", size = 3.5, fontface = "italic") +
  scale_x_date(date_breaks = "6 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = comma) +
  labs(
    title    = NULL,
    subtitle = "Bars = weekly count · line = 4-week rolling average",
    x        = NULL,
    y        = "Fatalities",
    caption  = "Source: ACLED"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", colour = "#1a2744"),
    plot.subtitle    = element_text(colour = "#666", size = 11),
    plot.caption     = element_text(colour = "#999", size = 9),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid.minor = element_blank()
  )
```

The timeline shows a dramatic escalation following the full-scale invasion in February 2022. 
This illustrates how a single geopolitical event can rapidly transform both the scale and intensity of violence.

# Pre- and post-2022 shift in global conflict intensity

```{r plot-conflict-shift}
library(dplyr)
library(ggplot2)
library(scales)
library(forcats)

shift_data <- combined |>
  filter(YEAR >= 2018, YEAR <= 2025) |>
  mutate(period = ifelse(WEEK < as.Date("2022-02-24"),
                         "Pre-2022",
                         "Post-2022")) |>
  group_by(BROAD_REGION, period) |>
  summarise(
    events = sum(EVENTS, na.rm = TRUE),
    fatalities = sum(FATALITIES, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(shift_data, aes(x = fct_reorder(BROAD_REGION, events), y = events, fill = period)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(
    values = c("Pre-2022" = "#95A5A6", "Post-2022" = "#C0392B"),
    name = NULL
  ) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Global Shift in Conflict Activity After 2022",
    subtitle = "Grey = pre-Ukraine war · Red = post-2022 escalation period",
    x = NULL,
    y = "Number of events",
    caption = "Source: ACLED"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "bottom"
  )
```


This chart compares conflict activity across regions before and after 2022. It highlights how the intensity of events shifted following 2022, with several regions experiencing a noticeable increase in violence and instability in the post-2022 period.
